Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-12 main.Rmd
software implementation added

It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore many possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

@@

@@

R package **gravitas** [@R-gravitas] is used to facilitate the systematic exploration here. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing we should have at our disposal is to know which all cyclic time granularities we can look at exhaustively. These cyclic granularities are a function of the index variable and are computed based on if they are circular (Section \ref{sec:circular-gran-def}), pseudo-circular (Section \ref{sec:quasi-circular-gran-def}) or aperiodic (Section \ref{sec:aperiodic-gran-def}). If we consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, fortnight, month, year), the following cyclic time granularities can be considered for this analysis. The interval of this tsibble is 30 minutes, and hence the temporal granularities may range from half-hour to year. The cyclic granularities which could be considered are "hhour_hour", "hhour_day", "hhour_week",  "hhour_fortnight",  "hhour_month", "hhour_year", "hour_day", "hour_week", "hour_fortnight", "hour_month", "hour_year", "day_week", "day_fortnight",  "day_month" , "day_year", "week_fortnight",  "week_month", "week_year" , "fortnight_month", "fortnight_month" and  "month_year". These options are too many and hence the most coarse temporal unit can be set to a “month” and some intermediate temporal units which might not be pertinent to the analysis like hhour or fortnight could be dropped. Then we will be left with six cyclic granularites namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month". Now that we have a list of cyclic granularities to look at, we should be able to compute them from the data using Sections \ref{sec:circular-gran-def}, \ref{sec:quasi-circular-gran-def} and \ref{sec:aperiodic-gran-def}.

```{r search, echo=FALSE, eval=FALSE}
library(gravitas)
smart_meter %>% search_gran(filter_out = c("semester", "quarter"))
```

@@

Now that we have a list of cyclic granularities to look at, we should be able to compute them from the data using Sections \ref{sec:circular-gran-def}, \ref{sec:quasi-circular-gran-def} and \ref{sec:aperiodic-gran-def}.

@@

Mumbai Indians (MI) and Chennai Super kings (CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into this question. Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble. From Figure \ref{fig:cricex}, it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that players are vulnerable to score through each over and make more runs in the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. To answer Q1, winning teams like CSK and MI have indeed different strategies when it comes to batting in the first or second innings.

@@

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. If a batsman is bowled out, it does not necessarily signify good fielding. So we only include number of catches and run out in an over as a measure of good fielding. Difference in runs across overs are likely to be negative if good fielding has an impact on the runs scored in the subsequent overs. Figure \ref{fig:field} shows that at least one dismissal in a previous over leads to a lower median and less spread in the distribution of run rates of the subsequent over.

@@

```{r field,  fig.cap = "Violin plots with an overlaid boxplot showing distribution of run rate across overs of the innings and (lagged) wickets dismissed due to fielding. This shows good fielding in previous over leads to lower run rate in the subsequent over. Also distribution of run rates has less spread in case there is at least one wicket in the previous over.", warning=FALSE}

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>% prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "violin",
) + ggtitle("") + geom_boxplot(width = 0.5, aes(colour = lag_field)) + ylim(0, 4.5)
```

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article.

>>>>

It is important to be able to navigate through all of these time granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore many possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

@@

The ideas presented in this article have been implemented in the open-source R [@R-language] package gravitas [@R-gravitas], available from CRAN. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.

@@

R package **gravitas** [@R-gravitas] is used to facilitate the systematic exploration here. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing we should have at our disposal is to know which all cyclic time granularities we can look at exhaustively. These cyclic granularities are a function of the index variable and are computed based on if they are circular <!--(Section \ref{sec:circular-gran-def})-->, pseudo-circular <!--(Section \ref{sec:quasi-circular-gran-def})--> or aperiodic <!--(Section \ref{sec:aperiodic-gran-def})-->. If we consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year), the following cyclic time granularities can be considered for this analysis. The interval of this tsibble is 30 minutes, and hence the temporal granularities may range from half-hour to year. The cyclic granularities which could be considered are "hhour_hour", "hhour_day", "hhour_week",  "hhour_month", "hhour_year", "hour_day", "hour_week", "hour_month", "hour_year", "day_week", "day_month" , "day_year",  "week_month", "week_year" and  "month_year". These options are too many and hence the most coarse temporal unit can be set to a “month” and some intermediate temporal units which might not be pertinent to the analysis like hhour could be dropped. Then we will be left with six cyclic granularites namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month". Now that we have a list of cyclic granularities to look at, we should be able to compute them from the data using Sections \ref{sec:circular-gran-def}, \ref{sec:quasi-circular-gran-def} and \ref{sec:aperiodic-gran-def}.

```{r search, echo=FALSE, eval = FALSE}
library(gravitas)
smart_meter %>% search_gran(filter_out = c("semester", "quarter", "fortnight"))
```

@@

<!-- Now that we have a list of cyclic granularities to look at, we should be able to compute them from the data using Sections \ref{sec:circular-gran-def}, \ref{sec:quasi-circular-gran-def} and \ref{sec:aperiodic-gran-def}. -->

@@

Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}, it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that players are vulnerable to score through each over and make more runs in the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. To answer Q1, winning teams like CSK and MI have indeed different strategies when it comes to batting in the first or second innings.

@@

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. <!--If a batsman is bowled out, it does not necessarily signify good fielding. So we only include number of catches and run out in an over as a measure of good fielding. 
Difference in runs across overs are likely to be negative if good fielding has an impact on the runs scored in the subsequent overs. --> Dismissals in previous over can lead to a batsman being either defensive or play more vulnerable. Figure \ref{fig:field} shows that no dismissals in the previous over will lead to a higher median and quartile spread of run rates compared to the case when there has been at least one dismissals.

@@

```{r field,  fig.cap = "Box plots showing distribution of run rate across overs of the innings and (lagged) wickets dismissed due to fielding. This shows at least one wicket in the previous over leads to downward shift of median run rate and quartile spread in the subsequent over for most overs.", warning = FALSE}

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>%
  filter(over!=1) %>% 
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)
) + ggtitle("") + scale_y_log10()
#+ geom_boxplot(width = 0.5, aes(colour = lag_field)) 
```

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Toscas [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/).
